---
title: 'California Survey Plot Data: Rosie'
author: "Brooke Rose"
date: "01/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r mac working directory, eval = FALSE}
# working directory
wd <- list()

# commonly used paths in my working directory
wd$data   <- "/Users/brookerose/Google Drive/Franklin_grant/project/data/"
wd$output <- "/Users/brookerose/Google Drive/Franklin_grant/project/output/"
wd$scripts <- "/Users/brookerose/Google Drive/Franklin_grant/project/scripts/"
```

```{r franklin working directory}
# working directory
wd <- list()

# commonly used paths in my working directory
wd$data   <- "/Users/Brooke Rose/Google Drive/Franklin_grant/project/data/"
wd$output <- "/Users/Brooke Rose/Google Drive/Franklin_grant/project/output/"
wd$scripts <- "/Users/Brooke Rose/Google Drive/Franklin_grant/project/scripts/"
wd$software <- "/Users/Brooke Rose/Google Drive/Franklin_grant/project/software/"
```

```{r loading libraries}
library("sf")
library("raster")
require(ggplot2)
require(flora) # for removing the species names from the authors' names
require(vroom) # read and write database
require(janitor) # for cleaning columns names (very useful for integrating databases later)
require(dplyr)
require(tidyr)
library(fuzzySim)
```

```{r extra data, echo = FALSE}
# Environmental data
env.stack <- stack(paste0(wd$output, 'predictors/env_stack/CFP_environmental_stack.grd')) # for coordinate reference system
cfp <- st_read(paste0(wd$data,"shapefiles/CFP/CFP_GIS.shp"))
cfp.trans <- st_transform(cfp, crs(env.stack))

us_states <- st_read(paste0(wd$data, "shapefiles/tl_2019_us_state/tl_2019_us_state.shp"))
california <- us_states %>%
  filter(NAME == "California")
```

# California Vegetation Survey Plots (from Rosie)
Received November 26, 2019. Includes a total of 25,449 plots

```{r Survey Plots}
survey_plots <- st_read(paste0(wd$data, 'BIOS/Cal_survey_points/Cal_survey_plots.shp'))
# Waypoint as character (for joining)
survey_plots$WyptID <- as.character(survey_plots$WyptID)
str(survey_plots)
#survey_plots <- survey_plots %>% janitor::clean_names()
```

# Plant List with Plot ID's (Waypoint)
List of plants at each plot, 461,718 "observations". Includes species name, code, species cover, and stratum (life form)

```{r plant list}
plant_list <- vroom::vroom(paste0(wd$data, 'BIOS/Cal_survey_points/SurveyPlants.csv'))
str(plant_list)
#plant_list <- plant_list %>% janitor::clean_names()
```

# Join survey plots and plant list
Creates a data frame with the waypoint ID, survey type, survey date, National Vegetation Classification Name, location information, species code, species name, and species cover

```{r joined dataframe}
survey_join <- left_join(survey_plots %>% dplyr::select(WyptID, SurveyType, SurveyDate, NVCS_Name,
                                                        geometry), 
                         plant_list %>% dplyr::select(WaypointID, CodeSpecies, SpeciesName, Species_cover),
                         by = c("WyptID" = "WaypointID"))

str(survey_join)

#st_write(survey_join, paste0(wd$data, "BIOS/Cal_survey_points/plot_plant_join.shp"), delete_layer = TRUE)
#write_csv(survey_join %>% dplyr::select(-geometry), paste0(wd$data, "BIOS/Cal_survey_points/plot_plant_join.csv"))
```


# Converting data to  presence absence matrix

```{r data cleaning}
survey_df <- vroom::vroom(paste0(wd$data, "BIOS/Cal_survey_points/plot_plant_join.csv"))
species_plots <- survey_df %>%
  dplyr::select(WyptID, CodeSpecies)

# converting plot/species data frame to species presence and absences
pres_abs <- splist2presabs(species_plots, sites.col = "WyptID", 
                           sp.col = "CodeSpecies", keep.n = FALSE)

survey_data <- pres_abs %>%
  right_join(survey_plots %>% dplyr::select(WyptID, geometry, SurveyType), by = c("WyptID" = "WyptID")) %>%
  st_as_sf() %>%
  dplyr::mutate(X = sf::st_coordinates(.)[,1],
                Y = sf::st_coordinates(.)[,2]) %>%
  st_set_geometry(NULL)

filt <- survey_data %>% dplyr::select(-c(1)) %>% colnames %>% sort
#filt #vector with species symbol

# Database with accepted names
plant_codes_accepted <-
  plant_list %>% dplyr::filter(CodeSpecies %in% filt)

rm(plant_list)
```

#### Removing incorrect coordinates
```{r santiago cleaning code 2}

### Remove plot without coordinates
cat("Total number of rapids plots: ", nrow(survey_data))
survey_data <- survey_data %>%
  dplyr::filter(complete.cases(survey_data %>% dplyr::select(X, Y)))
cat("Number of rapids plots with coordinates: ", nrow(survey_data))

### Remove plot with wrong coordinates
survey_data %>% ggplot(aes(X, Y)) + 
  geom_hex(bins=200) + 
  coord_equal() + theme_minimal()

### Remove columns with species without any presence
filt <- survey_data %>% dplyr::select(c(1)) %>% names
survey_data[2:6512] <- lapply(survey_data[2:6512], as.numeric)
pres_col <- colSums(survey_data %>% dplyr::select(-filt), na.rm = T)
pres_col <- pres_col[pres_col>0] %>% names

survey_data <- survey_data %>% dplyr::select(c(1:7), {pres_col}) 
survey_data <- survey_data %>% dplyr::select(-old_plot_id)

# nrow(survey_data)
cat("Number of species with at least one presence point: ", length(pres_col))

### Fill with zero in those columns with species names
survey_data <-
  survey_data %>% 
  dplyr::mutate(dplyr::across(pres_col, ~ tidyr::replace_na(.x, 0)))

survey_data <- survey_data %>% arrange(WyptID) #sort rows based on waypoint ID

### Remove  identical rows coordinates and summarize information at waypoint level 
survey_data %>% dplyr::count(WyptID) %>% arrange(desc(n)) #almost all waypoint is repeated

df2 <- # unique data for each plot
  survey_data %>% dplyr::select(-c(pres_col)) %>% unique

df3 <- survey_data %>% # summarizing species records by plot 
  dplyr::select(c('WyptID', c(pres_col))) %>%
  group_by(WyptID) %>%
  summarise(across(c(pres_col), ~ sum(.x)))

survey_data2 <- dplyr::left_join(df2, df3, by = "WyptID")
#dim(survey_data)
cat("Dimensions of clean data frame: ", dim(survey_data2))
survey_data <- survey_data2
rm(survey_data2)

nrow(unique(survey_data)) == nrow(survey_data) # all rows are unique. however there are plot with same coordinates 
# which not does not necessarily represent duplicate coordinates with different species composition

### Exploring plot without any species
rowfilt <- survey_data[pres_col] %>% colSums()
cat("All plots have at least one presence: ", range(rowfilt)) # all plots have presences but ZZZD has 78,033
```

```{r}
cat("Total number of unique plots in CFW dataset: ", length(unique(survey_data$WyptID)))
dups <- duplicated(survey_data[c('geometry')]) # remove duplicate coordinates
survey_uniq <- survey_data[!dups,] # plots with unique coordinates
cat(nrow(survey_data)-nrow(survey_uniq), "records removed with duplicate coordinates")
survey_uniq <- survey_uniq[, !duplicated(colnames(survey_uniq))]

```


# Map of all survey plots

```{r plotting}
# Writing to file as gz file
# So much faster than csv and shapefile
vroom::vroom_write(survey_uniq, paste0(wd$data, "BIOS/Cal_survey_points/pres_abs_matrix.gz"))
survey_df <- vroom::vroom(paste0(wd$data, "BIOS/Cal_survey_points/pres_abs_matrix.gz"))

#write_csv(survey_uniq, paste0(wd$data, "BIOS/Cal_survey_points/pres_abs_matrix.csv"))
#study_mat <- st_read(paste0(wd$data, "BIOS/Cal_survey_points/pres_abs_matrix.csv"))

study_sf <- st_as_sf(survey_df, coords = c("X", "Y"), crs = 3310)  
study_sf <- st_transform(study_sf, st_crs(cfp.trans)) # data frame of presence/absence for Compadre species with shapefile capabilities

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = study_sf, aes(color = SurveyType)) 

cat("Total number of plots: ", nrow(study_sf))
```

```{r data source}
cfw_sources <- study_sf %>%
  dplyr::mutate(survey = ifelse(SurveyType == 'Transect' | SurveyType == 'Releve' | 
                           SurveyType == 'Multivisit Releve' | SurveyType == 'releve' | 
                           SurveyType == 'Multivisit Transect' |
                           SurveyType == 'Multi-visit releve' | 
                           SurveyType == 'multi-visit transect' | 
                           SurveyType == 'Relevee', 'releve', 'rapid'),
                X = sf::st_coordinates(.)[,1],
                Y = sf::st_coordinates(.)[,2])

cfw_cfp <- st_intersection(cfw_sources, st_make_valid(cfp.trans))
cat("Total number of plots in the CFP: ", nrow(cfw_cfp))

vroom::vroom_write(cfw_cfp, paste0(wd$data, "BIOS/Cal_survey_points/pres_abs_matrix_cfp.gz"))
```

# Cal Fish & Wildlife Data by Survey Type
## Rapids

```{r rapids}
cfp_rapids <- cfw_cfp %>% filter(survey == 'rapid')
cat("Total number of rapid plots in the CFP: ", nrow(cfp_rapids))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = cfp_rapids, aes(color = SurveyType)) 
```

## Relevees

```{r relevees}
cfp_releve <- cfw_cfp %>% filter(survey == 'releve')
cat("Total number of relevee plots in the CFP: ", nrow(cfp_releve))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = cfp_releve, aes(color = SurveyType)) 


```

