---
title: "Thorne Rapids Data"
author: "Brooke Rose"
date: "1/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r libraries}
library("sf")
library("raster")
require(ggplot2)
require(flora) # for removing the species names from the authors' names
require(vroom) # read and write database
require(janitor) # for cleaning columns names (very useful for integrating databases later)
require(dplyr)
require(tidyr)
```

```{r mac working directory}
# Setting directories
# working directory
wd <- list()

# commonly used paths in my working directory
wd$data   <- "/Users/brookerose/Google Drive/Franklin_grant/project/data/"
wd$output <- "/Users/brookerose/Google Drive/Franklin_grant/project/output/"
```

```{r extra data, echo = FALSE}
# Environmental data
env.stack <- stack(paste0(wd$output, 'predictors/env_stack/CFP_environmental_stack.grd')) # for coordinate reference system
cfp <- st_read(paste0(wd$data,"shapefiles/CFP/CFP_GIS.shp"))
cfp.trans <- st_transform(cfp, crs(env.stack))

us_states <- st_read(paste0(wd$data, "shapefiles/tl_2019_us_state/tl_2019_us_state.shp"))
california <- us_states %>%
  filter(NAME == "California")
```

# Exploring Jim Thorne's rapid plot dataset

## USDA Plant Codes
```{r plant codes}
plant_codes <- vroom::vroom(paste0(wd$data, 'Thorne_plots/California USDA Plants codes.csv'))
str(plant_codes)
plant_codes <- plant_codes %>% janitor::clean_names()
```

## EPSG Code Descriptions
### 3309: NAD_1927_California_Teale_albers

## Rapids data
```{r rapids data}
#ac_rapids <- read_csv(paste0(wd$data, 'Thorne_plots/tbl_Rapid_A-C_GIS.csv')) # Species A-C
#cp_rapids <- read_csv(paste0(wd$data, 'Thorne_plots/tbl_Rapid_C-P_GIS.csv')) # Species C-P
#pz_rapids <- read_csv(paste0(wd$data, 'Thorne_plots/tbl_Rapid_P-Z_GIS.csv')) # Species P-Z

all_rapids <-
  list.files(paste0(wd$data, 'Thorne_plots'),
             pattern = 'tbl_R',
             full.names = T)  %>%
  sapply(., vroom::vroom)
```

### Cleaning data
```{r santiago cleaning code 1}
all_rapids <- dplyr::bind_rows(all_rapids) 

filt <- all_rapids %>% dplyr::select(-c(1:8)) %>% colnames %>% sort
#filt #vector with species symbol

# Database with accepted names
plant_codes_accepted <-
  plant_codes %>% dplyr::filter(symbol %in% filt)

# Database with accepted and synonyms names for those columns names with synonyms
plant_codes_synonym <-
  plant_codes %>% dplyr::filter(synonym_symbol %in% filt)
plant_codes_synonym <-
  plant_codes %>% dplyr::filter(symbol %in% plant_codes_synonym$symbol)
plant_codes_synonym <-
  plant_codes_synonym %>% filter(synonym_symbol %in% filt |
                                   is.na(synonym_symbol))

rm(plant_codes)
```

#### Removing incorrect coordinates
```{r santiago cleaning code 2}

### Clean columns names 
filt <- all_rapids %>% dplyr::select(c(1:8)) %>% names
newnames <- all_rapids[1, filt] %>% 
  janitor::clean_names() %>% names
colnames(all_rapids)[1:8] <- newnames

### Remove plot without coordinates
cat("Total number of rapids plots: ", nrow(all_rapids))
all_rapids <- all_rapids %>%
  dplyr::filter(complete.cases(all_rapids %>% dplyr::select(x_final, y_final)))
cat("Number of rapids plots with coordinates: ", nrow(all_rapids))

### Remove plot with wrong coordinates
all_rapids %>% ggplot(aes(x_final, y_final)) + 
  geom_hex(bins=200) + 
  coord_equal() + theme_minimal()
all_rapids <- all_rapids %>% dplyr::filter(x_final<600000 & x_final >-500000)
all_rapids %>% ggplot(aes(x_final, y_final)) + 
  geom_hex(bins=200) + 
  coord_equal() + theme_minimal()

### Remove columns with species without any presence
filt <- all_rapids %>% dplyr::select(c(1:8)) %>% names
pres_col <- colSums(all_rapids %>% dplyr::select(-filt), na.rm = T)
pres_col <- pres_col[pres_col>0] %>% names

all_rapids <- all_rapids %>% dplyr::select(c(1:8), {pres_col}) 
all_rapids <- all_rapids %>% dplyr::select(-old_plot_id, -elev_orig)

# nrow(all_rapids)
cat("Number of species with at least one presence point: ", length(pres_col))

### Fill with zero in those columns with species names
all_rapids <-
  all_rapids %>% 
  dplyr::mutate(dplyr::across(pres_col, ~ tidyr::replace_na(.x, 0)))

all_rapids <- all_rapids %>% arrange(new_id) #sort rows based on new_id

### Remove  identical rows coordinates and summarize information at new_id level 
all_rapids %>% dplyr::count(new_id) %>% arrange(desc(n)) #almost all new_id is repeated

df2 <- # unique data for each plot
  all_rapids %>% dplyr::select(-c('total', pres_col)) %>% unique

df3 <- all_rapids %>% # summarizing species records by plot 
  dplyr::select(c('new_id', c('total', pres_col))) %>%
  group_by(new_id) %>%
  summarise(across(c('total', pres_col), ~ sum(.x)))

all_rapids2 <- dplyr::left_join(df2, df3, by = "new_id")
#dim(all_rapids)
cat("Dimensions of clean data frame: ", dim(all_rapids2))
all_rapids <- all_rapids2
rm(all_rapids2)

# nrow(unique(all_rapids)) == nrow(all_rapids) # all rows are unique. however there are plot with same coordinates 
# which not does not necessarily represent duplicate coordinates with different species composition

### Exploring plot without any species
rowfilt <- all_rapids[pres_col] %>% colSums()
cat("All plots have at least one presence: ", range(rowfilt)) # all plots have presences 

# some plots
all_rapids %>% ggplot(aes(x_final, y_final)) + 
  geom_hex(bins=200) + 
  coord_equal() + theme_minimal()

# Save a cleaned database
vroom::vroom_write(all_rapids, paste0(wd$data, "Thorne_plots/cleaned_rapid_study_species.gz"))
```

```{r data with species names}

#### Database with species names ####
all_rapids <- vroom::vroom(paste0(wd$data, "Thorne_plots/cleaned_rapid_study_species.gz"))
filt <- names(all_rapids)[1:6]
pres_col <- names(all_rapids)[-c(1:6)]

for( i in 1:length(pres_col)){
  message('processing species name ', i)
  
  sp_symbol <- pres_col[i]
  if(sp_symbol %in% plant_codes_accepted$symbol) {
    sp_names <- plant_codes_accepted %>%
      dplyr::filter(symbol %in% sp_symbol) %>%
      pull(scientific_name_with_author)
    sp_names <- flora::remove.authors(sp_names)
    pres_col2 <- pres_col[i]
    names(pres_col2) <- sp_names
    all_rapids <- dplyr::rename(all_rapids, pres_col2)

  } else if (sp_symbol %in% plant_codes_synonym$synonym_symbol) {
    message('Updating species names')
    sp_symbol2 <- plant_codes_synonym %>%
      dplyr::filter(synonym_symbol %in% sp_symbol) %>% #note here is used synonym_symbol
      pull(symbol) #here is extracted the accepted names symbol
    sp_names <- plant_codes_synonym %>%
      dplyr::filter(symbol %in% sp_symbol2) %>%
      dplyr::filter(is.na(synonym_symbol)) %>%
      pull(scientific_name_with_author)

    sp_names <- flora::remove.authors(sp_names)
    pres_col2 <- pres_col[i]
    names(pres_col2) <- sp_names
    all_rapids <- dplyr::rename(all_rapids, pres_col2)
    
  } else {
    message('This species code has no species names in USDA database')
  }
}

vroom::vroom_write(all_rapids, 
                   paste0(wd$data, "Thorne_plots/Thorne_rapids_cleaned_w_sp_names.gz"))

```


## Study Species Codes
```{r species code list}
sp.list <- c('WyptID', 'ABMA', 'ABMAS2', 'ABMAS', 'ABMAC', 'ABBR', 'ACIL',
             'ACMI2', 'ACMIA', 'ACMIF', 'ACMIA2', 'ACMIM4', 'ACMIC', 'ACMIM2','ACMIO', 'ACMIL', 
             'ACMIO2', 'ACMIP5', 'ACMIA4', 'ACMIG2', 'ACMIL3', 'ACMIR', 'ACMIR2',
             'ADSP', 'AECA', 'ARCA33', 'ARGLC', 'ARGL3', 'ARGLA2', 'ARGLA', 'ARGLC4', 
             'ARGLC2', 'ARGLG3', 'ARGLC3', 'ARGLG4', 'ARGLM2', 'ARGLM', 'ARGLZ2', 
             'ARGLH2', 'ARGLH', 'ARGLZ', 'ARRA7', 'ARRU2', 'ARCAB3', 'ARRO4', 'CADE27', 'CAAL2', 
             'CAALR', 'CAPU2', 'CAOB', 'CATI3', 'CEGRP', 'CEGRP2', 'CEPE6', 'CEGR', 'CEGRV2', 'CEME', 
             'CEMEI', 'CEMEI2', 'CEMEM2', 'CEMEP', 'CEOL', 'CEOLO5', 'CEOLO', 'CESO2', 
             'CEOLS', 'CEOLS2', 'CETO', 'CETOO2', 'CETOO', 'CEVE2', 'CHFO', 'CHOR', 'COCA23',
             'COCOC', 'COCA23', 'COCOC9', 'DECO13', 'DEHAH', 'DEHEH2', 'DEHE', 'DEHEC', 'DEHEH', 
             'DEHEP', 'DEPAP5', 'DEHES2', 'DERE2', 'DEHER', 'ERER11', 'ERPE5', 'ERPEP4', 'ERPES2', 
             'ERPEV', 'ERAR11', 'ERARA', 'ERARH', 'ERARP', 'ERARP2', 'GAAN2', 'GAANA2', 
             'GAANB2', 'GAAND', 'GAANS', 'GAANT', 'GAANB', 'GAANF', 'GAANF2', 'GAANG2',
             'GAANG3', 'GAANJ', 'GAANN', 'GAANO', 'GAANO2', 'GAJE', 'GAANS2', 'GAJO', 
             'GAANP2', 'HASQ2', 'HASQG2', 'HASQO2', 'HASQS2', 'HEFO10', 'CUFO2', 'HESA17', 
             'HEST10', 'HEAR5', 'HEARA2', 'HEARC2', 'HEARM', 'JUCA', 'JUCAC5', 'JUCAQ', 
             'JUCAQ2', 'JUHI', 'JUCAH', 'KEAN', 'KEANA', 'KEANA3', 'KEANM', 'KEANM2', 
             'LECA3', 'LUAR', 'LUEX2', 'LUARE', 'LUTI', 'LUTIL', 'LUTIT', 'MICA3', 'MICAE', 
             'MICAG', 'MICAR', 'MICAV', 'ERCA23', 'MIGL2', 'MIGUG', 'MIGU', 'MIGUA', 
             'MIGUA2', 'MIGUH', 'MIGUL2', 'MIGUM', 'MIGUS', 'MIGUA3', 'MIGUD', 'MIGUD2', 
             'MIGUG3', 'MIGUG2', 'MIGUH2', 'MIGUI', 'MIGUL4', 'MIGUL3', 'MIGUM2', 'MIGUN', 'MIGUP', 
             'MILE2', 'MIOB2', 'PHIN13', 'PHINC2', 'PHINI4', 'PIBR', 'PIMO5', 'PIMOM', 'PIMOT2',
             'PIMOT', 'PIAT', 'PIBA', 'PIPAA', 'PIBAB', 'PICO3', 'PIJE', 'PILA', 'PIMU', 
             'PIMUB2', 'PIMUS2', 'PIMUB', 'PIMUM', 'PIMUR2', 'PIMUS', 'PISA2', 'POST11', 
             'ARAN7', 'POAN5', 'POANA2', 'POANC', 'POANS2', 'POANY', 'POANE', 'POANP', 
             'POAANG', 'POANL2', 'POANR2', 'PSMA', 'PTKI', 'STKI', 'QUDO', 'QUDU', 
             'QUDUD', 'QUDUE', 'QUDUM2', 'QUDUT', 'QUAL2', 'QUDUA2', 'QUDUK', 'QUMA4', 'QUEN', 
             'QULO', 'QULOA', 'QULOI', 'QULOT', 'QULOW', 'QUSA2', 'QUWI2', 'QUWIF', 
             'QUWIW', 'FRRUM', 'RHRUM', 'RHRUO', 'RHRU', 'RHRUY', 'RILA2', 'SCCA3', 'LECA3',
             'RILA2', 'RIMA', 'RIMAM', 'RIMAC', 'RIMAV', 'SALE3', 'SESE3', 'TOCA', 'TRLA3',
             'UMCAC', 'UMCA', 'UMCAF', 'VILA3', 'XYBI')
```

## Rapids Data
### What are the data sources for the rapid plots?
```{r data sources}
rapids_sarea <- c(unique(all_rapids$sarea))
print(rapids_sarea)
```

### Number of rapid plots by source
```{r notes on data sources}
rapid_source <- all_rapids %>% group_by(sarea) %>% summarise(count=n())
rapid_source
```

### Converting plot data to Spatial Feature 
#### These data use EPSG 3309 (NAD_1927_California_Teale_albers)

```{r rapids plots}
rapids_sf <- st_as_sf(all_rapids %>% filter(sarea != 'PCTM'), coords = c('x_final', 'y_final'), crs = 3309) # converting to spatial object
rapids_trans <- st_transform(rapids_sf, crs(env.stack))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = rapids_trans, aes(color = sarea))
```

### Plots with sketchy coordinates
#### PCT: These seem to follow the Pacific Crest Trail, maybe pretty biased? 

```{r PCTM plots}
pctm_sf <- st_as_sf(all_rapids %>% filter(sarea == 'PCTM'), coords = c('x_final', 'y_final'), crs = 3309) # converting to spatial object
pctm_trans <- st_transform(pctm_sf, crs(env.stack))
cat("Number of PCTM plots total: ", nrow(pctm_trans))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = pctm_trans, aes(color = sarea))

pctm_cfp <- st_intersection(pctm_trans, st_make_valid(cfp.trans)) # selecting only PCTM in CFP
cat("Number of PCTM plots in CFP: ", nrow(pctm_cfp))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = pctm_cfp, aes(color = sarea))
```

#### CNPS: California Native Plant Society
##### 6 plots have "off" coordinates, these plots also use EPSG 3309 (NAD_1927_California_Teale_albers)

```{r CNPS plots}
cnps_sf <- st_as_sf(all_rapids %>% filter(sarea == 'CNPS'), coords = c('x_final', 'y_final'), crs = 3309) # converting to spatial object
cnps_trans <- st_transform(cnps_sf, crs(env.stack))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = cnps_trans, aes(color = sarea))

cnps_cfp <- st_intersection(cnps_trans, cfp.trans) # selecting only CNPS in CFP
ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = cnps_cfp, aes(color = sarea))
```

#### CARZ

```{r CARZ plots}
carz_sf <- st_as_sf(all_rapids %>% filter(sarea == 'CARZ'), coords = c('x_final', 'y_final'), crs = 3309) # converting to spatial object
carz_trans <- st_transform(carz_sf, crs(env.stack))
cat("Number of CARZ plots total: ", nrow(carz_trans))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = carz_trans, aes(color = sarea))

carz_cfp <- st_intersection(carz_trans, cfp.trans) # selecting only CNPS in CFP
cat("Number of CARZ plots in CFP: ", nrow(carz_cfp))
```


### Integrating the plots
The final data include our study species with plots that are located within the California Floristic Province. 

```{r integrating data}
rapids2_sf <- st_as_sf(all_rapids, coords = c('x_final', 'y_final'), crs = 3309) # converting to spatial object
rapids2_trans <- st_transform(rapids2_sf, crs(env.stack))
rapids_cfp <- st_intersection(rapids2_trans, st_make_valid(cfp.trans))
cat("Number of rapid plots in CFP: ", nrow(rapids_cfp))

ggplot() +
  geom_sf(data = cfp.trans) +
  geom_sf(data = rapids_cfp, aes(color = sarea))

rapids_sp <- rapids_cfp %>%
  dplyr::mutate(X = sf::st_coordinates(.)[,1],
                Y = sf::st_coordinates(.)[,2])


vroom::vroom_write(rapids_sp, paste0(wd$data, "Thorne_plots/final_compiled_rapid.gz"))
```